---
title: "Thesis Model: MSTE-JM (Method 7)"
author: "Tiffany"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

# Overview

This notebook runs **Method 7: MSTE-JM** — the thesis model combining a
multimodal autoencoder (MAE) with a joint model (JM) of longitudinal MMSE
trajectories and time-to-MCI-conversion.

**Input files required:**

- `latent_patient_level_v6.csv` — patient-level features and survival labels
- `latent_improved_autoencoder_v6.csv` — longitudinal visit-level data

**Output:** Figures saved to `thesis_figures/MSTE-JM/`

---

# 1. Setup

## 1.1 Packages

```{r packages}
required_packages <- c(
  "survRM2", "boot", "JMbayes2", "dplyr", "nlme",
  "survival", "pec", "survminer", "ggplot2", "gridExtra",
  "tidyr", "glmnet", "e1071", "scales"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.r-project.org")
    library(pkg, character.only = TRUE)
  }
}

# Avoid namespace conflicts
select    <- dplyr::select
summarize <- dplyr::summarize

# Output directory
if (!dir.exists("thesis_figures")) dir.create("thesis_figures")

cat("Setup complete.\n")
```

---

# 2. Helper Functions

These functions are shared utilities used throughout the analysis.

## 2.1 Feature Selection & Standardisation

```{r helpers-features}
# Guard against glmnet failing on very small samples
safe_cv_glmnet <- function(x_mat, y_mat, alpha = 0.5) {
  n <- nrow(x_mat)
  if (n < 5) {
    cat(sprintf("  WARNING: Only %d patients — skipping glmnet\n", n))
    return(NULL)
  }
  nfolds <- min(10L, n)
  if (nfolds < 10)
    cat(sprintf("  WARNING: Small N (%d): nfolds=%d\n", n, nfolds))
  tryCatch(
    cv.glmnet(x_mat, y_mat, family = "cox", alpha = alpha,
              nfolds = nfolds, type.measure = "C"),
    error = function(e) {
      cat(sprintf("  WARNING: cv.glmnet failed (%s)\n", e$message)); NULL
    }
  )
}

# Limit features to Events-Per-Event (EPE) threshold of 10
cap_features_by_epe <- function(sel, events, epe = 10) {
  sel    <- as.character(sel)
  events <- as.integer(events)
  max_f  <- max(1L, floor(events / epe))
  if (length(sel) > max_f) {
    cat(sprintf("  EPE cap: %d events -> max %d features (was %d)\n",
                events, max_f, length(sel)))
    sel <- sel[seq_len(max_f)]
  }
  sel
}

# Z-score standardise features using training set statistics
standardise_features <- function(train_df, test_df, feat_cols) {
  feat_cols <- intersect(feat_cols, intersect(names(train_df), names(test_df)))
  for (f in feat_cols) {
    train_df[[f]] <- as.numeric(train_df[[f]])
    test_df[[f]]  <- as.numeric(test_df[[f]])
    m <- mean(train_df[[f]], na.rm = TRUE)
    s <- sd(train_df[[f]],   na.rm = TRUE)
    if (!is.na(s) && s > 1e-10) {
      train_df[[f]] <- pmin(pmax((train_df[[f]] - m) / s, -5), 5)
      test_df[[f]]  <- pmin(pmax((test_df[[f]]  - m) / s, -5), 5)
    } else {
      train_df[[f]] <- 0.0
      test_df[[f]]  <- 0.0
    }
  }
  list(train = train_df, test = test_df)
}

# Select features via elastic-net, falling back to top univariate C-index
select_features <- function(train_df, all_cols, n_events, method_name = "") {
  all_cols <- intersect(all_cols, names(train_df))
  ok_var   <- sapply(all_cols, function(f) {
    v <- var(as.numeric(train_df[[f]]), na.rm = TRUE)
    !is.na(v) && v > 1e-10
  })
  all_cols <- all_cols[ok_var]

  if (length(all_cols) == 0) {
    cat(sprintf("  WARNING: %s: no valid columns\n", method_name))
    return(character(0))
  }

  x_mat  <- as.matrix(train_df[, all_cols, drop = FALSE])
  y_mat  <- Surv(train_df$time, train_df$event)
  cv_fit <- safe_cv_glmnet(x_mat, y_mat)

  sel <- character(0)
  if (!is.null(cv_fit)) {
    coefs <- coef(cv_fit, s = cv_fit$lambda.1se)
    sel   <- all_cols[which(coefs != 0)]
    if (length(sel) < 5) {
      coefs <- coef(cv_fit, s = cv_fit$lambda.min)
      sel   <- all_cols[which(coefs != 0)]
    }
  }

  if (length(sel) < 5) {
    cat(sprintf("  WARNING: %s: glmnet gave <5 features — top-10 univariate\n", method_name))
    uni_c <- sapply(all_cols, function(f) tryCatch(
      summary(coxph(as.formula(paste("Surv(time,event) ~", f)),
                    data = train_df))$concordance[1],
      error = function(e) 0.5
    ))
    top_idx <- order(uni_c, decreasing = TRUE)[seq_len(min(10L, length(all_cols)))]
    sel     <- all_cols[top_idx]
  }

  sel <- cap_features_by_epe(sel, n_events)
  cat(sprintf("  Selected %d features\n", length(sel)))

  # Sanity check: drop features that cause infinite Cox coefficients
  if (length(sel) > 1) {
    test_cox <- tryCatch(
      coxph(as.formula(paste("Surv(time,event) ~", paste(sel, collapse=" + "))),
            data = train_df, x = FALSE,
            control = coxph.control(iter.max = 30)),
      error = function(e) NULL
    )
    if (!is.null(test_cox)) {
      cv <- tryCatch(coef(test_cox), error = function(e) NULL)
      if (!is.null(cv) && any(!is.finite(cv))) {
        cat(sprintf("  WARNING: Infinite coefficients — falling back to top-1\n"))
        uni_c2  <- sapply(sel, function(f) tryCatch(
          summary(coxph(as.formula(paste("Surv(time,event) ~", f)),
                        data = train_df))$concordance[1],
          error = function(e) 0.5
        ))
        sel <- sel[which.max(uni_c2)]
      }
    }
  }
  sel
}
```

## 2.2 Cox & Survival Utilities

```{r helpers-cox}
# Ridge Cox regression with cross-validated lambda
fit_ridge_cox <- function(train_df, sel, method_name = "") {
  x_mat  <- as.matrix(train_df[, sel, drop = FALSE])
  y_mat  <- Surv(train_df$time, train_df$event)
  nfolds <- min(10L, nrow(x_mat))
  lambda_grid <- exp(seq(log(0.001), log(10), length.out = 100))

  cv_ridge <- tryCatch(
    cv.glmnet(x_mat, y_mat, family = "cox", alpha = 0,
              lambda = lambda_grid, nfolds = nfolds, type.measure = "C"),
    error = function(e) {
      cat(sprintf("  WARNING: %s: ridge cv failed — unpenalised fallback\n", method_name))
      NULL
    }
  )

  if (is.null(cv_ridge)) {
    return(coxph(as.formula(paste("Surv(time,event) ~", paste(sel, collapse=" + "))),
                 data = train_df, x = TRUE, method = "breslow"))
  }

  ridge_coefs <- as.numeric(coef(cv_ridge, s = cv_ridge$lambda.min))
  names(ridge_coefs) <- sel

  if (max(abs(ridge_coefs)) < 0.001) {
    best_lambda <- cv_ridge$lambda[which.max(cv_ridge$cvm)]
    ridge_coefs <- as.numeric(coef(cv_ridge, s = best_lambda))
    names(ridge_coefs) <- sel
  }

  tryCatch(
    coxph(as.formula(paste("Surv(time,event) ~", paste(sel, collapse=" + "))),
          data = train_df, x = TRUE, method = "breslow",
          init = ridge_coefs, control = coxph.control(iter.max = 0)),
    error = function(e) {
      coxph(as.formula(paste("Surv(time,event) ~", paste(sel, collapse=" + "))),
            data = train_df, x = TRUE, method = "breslow")
    }
  )
}

# Median survival time: capped at 95th percentile of observed times
get_median_survival <- function(coxFit, newdata) {
  bh      <- basehaz(coxFit, centered = FALSE)
  lp      <- predict(coxFit, newdata = newdata, type = "lp")
  obs_cap <- quantile(newdata$time, 0.95, na.rm = TRUE)

  sapply(lp, function(l) {
    surv_i <- exp(-bh$hazard * exp(l))
    idx    <- which(surv_i <= 0.5)
    if (length(idx) > 0) min(bh$time[min(idx)], obs_cap)
    else min(bh$time[which.min(surv_i)], obs_cap)
  })
}
```

## 2.3 LME and Joint Model Fitting

```{r helpers-jm-fit}
# LME with random slope; falls back to random intercept on convergence failure
safe_lme <- function(formula, data, random_slope = TRUE) {
  ctrl <- lmeControl(opt = "optim", maxIter = 500, returnObject = FALSE)
  if (random_slope) {
    tryCatch(
      lme(formula, random = ~Years_bl | PTID, data = data, control = ctrl),
      error = function(e) {
        cat(sprintf("  WARNING: Random-slope LME failed (%s) — intercept-only\n",
                    trimws(e$message)))
        tryCatch(
          lme(formula, random = ~1 | PTID, data = data, control = ctrl),
          error = function(e2) stop(sprintf("LME intercept-only failed: %s", e2$message))
        )
      }
    )
  } else {
    lme(formula, random = ~1 | PTID, data = data, control = ctrl)
  }
}

# Joint model: value+slope association first, then value-only fallback
safe_jm <- function(cox_fit, lme_fit, time_var = "Years_bl", seed = 42,
                    method_name = "") {
  cat("  Fitting JM (value+slope, 10000 iter, 2 chains)...\n")
  jm1 <- tryCatch(
    jm(cox_fit, lme_fit, time_var = time_var,
       functional_forms = ~value(MMSE) + slope(MMSE),
       n_iter = 10000, n_burnin = 2000, n_thin = 5, n_chains = 2, seed = seed),
    error = function(e) {
      cat(sprintf("  WARNING: JM (value+slope) failed (%s) — trying value-only\n",
                  trimws(e$message)))
      NULL
    }
  )
  if (!is.null(jm1)) { cat("  JM (value+slope) converged\n"); return(jm1) }

  cat("  Fitting JM (value-only, 7000 iter, 2 chains)...\n")
  jm2 <- tryCatch(
    jm(cox_fit, lme_fit, time_var = time_var,
       functional_forms = ~value(MMSE),
       n_iter = 7000, n_burnin = 1500, n_thin = 5, n_chains = 2, seed = seed),
    error = function(e) {
      cat(sprintf("  WARNING: JM (value-only) also failed (%s)\n", trimws(e$message)))
      NULL
    }
  )
  if (!is.null(jm2)) cat("  JM (value-only) converged\n")
  jm2
}
```

## 2.4 Joint Model Evaluation Helpers

```{r helpers-jm-eval}
# Extract association parameters (alpha) from the JM posterior summary
extract_jm_association <- function(jointFit, method_dir, name) {
  res <- list(
    jm_association_val         = NA, jm_association_val_lower   = NA,
    jm_association_val_upper   = NA, jm_association_slope       = NA,
    jm_association_slope_lower = NA, jm_association_slope_upper = NA
  )
  if (is.null(jointFit)) return(res)

  sm  <- tryCatch(summary(jointFit), error = function(e) NULL)
  if (is.null(sm)) return(res)
  tbl <- tryCatch(as.data.frame(sm$Survival), error = function(e) NULL)
  if (is.null(tbl)) return(res)

  cn        <- colnames(tbl)
  mean_col  <- cn[grep("^[Mm]ean$|^[Ee]stimate$", cn)][1]
  lower_col <- cn[grep("2\\.5|[Ll]ower", cn)][1]
  upper_col <- cn[grep("97\\.5|[Uu]pper", cn)][1]
  rn        <- rownames(tbl)

  vi <- grep("value|Assoct$", rn, ignore.case = TRUE)
  if (length(vi) > 0) {
    res$jm_association_val       <- as.numeric(tbl[vi[1], mean_col])
    res$jm_association_val_lower <- as.numeric(tbl[vi[1], lower_col])
    res$jm_association_val_upper <- as.numeric(tbl[vi[1], upper_col])
  }
  si <- grep("slope|Assoct_slope", rn, ignore.case = TRUE)
  if (length(si) > 0) {
    res$jm_association_slope       <- as.numeric(tbl[si[1], mean_col])
    res$jm_association_slope_lower <- as.numeric(tbl[si[1], lower_col])
    res$jm_association_slope_upper <- as.numeric(tbl[si[1], upper_col])
  }

  write.csv(tbl,
            file.path(method_dir, paste0(gsub(" ", "_", name), "_jm_association.csv")),
            row.names = TRUE)
  res
}

# Dynamic AUC at landmark time points using tvAUC()
compute_dynamic_auc <- function(jointFit, long_data, surv_data,
                                patient_level_data, cox_vars,
                                method_dir, name, min_events_for_auc = 4) {
  res_auc <- NA
  if (is.null(jointFit) || is.null(long_data) || nrow(long_data) == 0) return(res_auc)

  long_data$PTID     <- as.character(long_data$PTID)
  long_data$Years_bl <- as.numeric(long_data$Years_bl)
  long_data$MMSE     <- as.numeric(long_data$MMSE)
  surv_data$PTID     <- as.character(surv_data$PTID)

  max_obs   <- max(surv_data$time, na.rm = TRUE)
  landmarks <- c(0.5, 1, 1.5, 2)
  landmarks <- landmarks[landmarks < max_obs * 0.80]

  cox_vars_detected <- tryCatch({
    f <- jointFit$model_info$terms_Surv_noResp
    if (!is.null(f)) setdiff(all.vars(f), c("time","event","PTID","Years_bl","MMSE"))
    else character(0)
  }, error = function(e) character(0))
  if (length(cox_vars_detected) == 0) cox_vars_detected <- cox_vars

  surv_outcome   <- surv_data %>% select(PTID, time, event)
  aucs           <- c()
  used_landmarks <- c()

  for (t0 in landmarks) {
    tryCatch({
      at_risk_ids <- surv_outcome$PTID[surv_outcome$time > t0]
      if (length(at_risk_ids) < 5) next

      long_t0 <- long_data %>%
        filter(PTID %in% at_risk_ids, Years_bl <= t0 + 1e-6) %>%
        left_join(surv_outcome, by = "PTID")

      if (length(cox_vars_detected) > 0 && !is.null(patient_level_data)) {
        missing_cox <- setdiff(cox_vars_detected, names(long_t0))
        if (length(missing_cox) > 0) {
          avail <- intersect(c("PTID", missing_cox), names(patient_level_data))
          if (length(avail) > 1) {
            long_t0 <- long_t0 %>%
              left_join(
                patient_level_data %>% select(all_of(avail)) %>%
                  distinct(PTID, .keep_all = TRUE),
                by = "PTID"
              )
          }
        }
      }

      long_t0 <- long_t0[complete.cases(
        long_t0[, intersect(c("PTID","Years_bl","MMSE","time","event"), names(long_t0))]), ]

      n_pts_t0 <- length(unique(long_t0$PTID))
      n_evt_t0 <- sum(surv_outcome$event == 1 & surv_outcome$time > t0 &
                        surv_outcome$time <= (t0 + 1.5) &
                        surv_outcome$PTID %in% unique(long_t0$PTID))

      if (n_pts_t0 < 5 || n_evt_t0 < min_events_for_auc) next

      auc_obj        <- tvAUC(jointFit, newdata = long_t0, Tstart = t0, Dt = 1.5)
      aucs           <- c(aucs, auc_obj$auc)
      used_landmarks <- c(used_landmarks, t0)
      cat(sprintf("    Dynamic AUC at t=%.1f: %.4f\n", t0, auc_obj$auc))

    }, error = function(e)
      cat(sprintf("    WARNING: tvAUC at t=%.1f failed: %s\n", t0, trimws(e$message))))
  }

  if (length(aucs) > 0) {
    res_auc <- mean(aucs, na.rm = TRUE)
    write.csv(data.frame(Landmark = used_landmarks, AUC = aucs),
              file.path(method_dir, paste0(gsub(" ","_",name), "_dynamic_auc.csv")),
              row.names = FALSE)
  }
  res_auc
}
```

## 2.5 Survival Curve Plotting

> **Thesis figure:** `MSTE-JM_jm_survival_curves.png`

This function produces two plots:
- **(A)** Individual predicted survival curves for a sample of converters and non-converters
- **(B)** Group-averaged survival curves aligned to a common time-since-landmark axis

```{r helpers-survplot}
plot_joint_model_survival_curves <- function(jointFit, surv_data, long_data,
                                             method_dir, name,
                                             n_individual = 12,
                                             pred_horizon = NULL) {
  if (is.null(jointFit) || is.null(long_data) || nrow(long_data) == 0) {
    cat("    WARNING: skipping survival curves\n"); return(invisible(NULL))
  }

  surv_data$PTID <- as.character(surv_data$PTID)
  long_data$PTID <- as.character(long_data$PTID)

  if (is.null(pred_horizon))
    pred_horizon <- quantile(surv_data$time, 0.90, na.rm = TRUE)

  # Helpers -------------------------------------------------------------------

  extract_pred_cols <- function(sm) {
    cn_lower <- tolower(names(sm))
    find_col <- function(patterns) {
      for (p in patterns) { idx <- grep(p, cn_lower); if (length(idx) > 0) return(names(sm)[idx[1]]) }
      NA_character_
    }
    list(
      time  = find_col(c("^times$","^time$")),
      surv  = find_col(c("pred_cif","pred_surv","^pred$","^mean$","^surv$")),
      lower = find_col(c("low_cif","low_surv","^low$","lower","2\\.5")),
      upper = find_col(c("upp_cif","upp_surv","^upp$","upper","97\\.5"))
    )
  }

  maybe_flip_to_survival <- function(df, grp_label = "") {
    ordered_df <- df[order(df$time), ]
    is_cif <- mean(diff(ordered_df$surv) > 0, na.rm = TRUE) > 0.6
    if (is_cif) {
      lo <- df$lower; up <- df$upper
      df$surv  <- 1 - df$surv
      df$lower <- 1 - up
      df$upper <- 1 - lo
    }
    df
  }

  prepare_jm_nd <- function(pt_long, surv_time) {
    t_last <- max(pt_long$Years_bl, na.rm = TRUE)
    pt_long$time  <- max(surv_time, t_last + 1e-4)
    pt_long$event <- 0L
    pt_long
  }

  pred_times_for <- function(t_last) {
    start <- t_last + 0.01
    if (start >= pred_horizon) return(seq(start, start + 0.5, length.out = 30))
    seq(start, pred_horizon, length.out = max(30L, round(60 * (pred_horizon - start) / pred_horizon)))
  }

  # (A) Individual curves -----------------------------------------------------
  eligible_ids   <- intersect(surv_data$PTID, unique(long_data$PTID))
  converters     <- surv_data$PTID[surv_data$event == 1 & surv_data$PTID %in% eligible_ids]
  non_converters <- surv_data$PTID[surv_data$event == 0 & surv_data$PTID %in% eligible_ids]
  n_each         <- max(1L, floor(n_individual / 2))
  set.seed(42)
  sampled_ids <- unique(c(
    sample(converters,     min(n_each, length(converters)),     replace = FALSE),
    sample(non_converters, min(n_each, length(non_converters)), replace = FALSE)
  ))

  curve_list <- list()
  for (pid in sampled_ids) {
    tryCatch({
      pt_long  <- long_data %>% filter(PTID == pid) %>% arrange(Years_bl)
      surv_row <- surv_data[surv_data$PTID == pid, , drop = FALSE]
      if (nrow(pt_long) < 1 || nrow(surv_row) == 0) next

      pt_long_jm <- prepare_jm_nd(pt_long, surv_row$time[1])
      sjm        <- predict(jointFit, newdata = pt_long_jm, process = "event",
                            times = pred_times_for(max(pt_long$Years_bl, na.rm=TRUE)),
                            return_newdata = TRUE)
      sm   <- as.data.frame(sjm)
      cols <- extract_pred_cols(sm)
      if (any(is.na(unlist(cols)))) next

      surv_est <- data.frame(time=sm[[cols$time]], surv=sm[[cols$surv]],
                             lower=sm[[cols$lower]], upper=sm[[cols$upper]])
      surv_est <- surv_est[is.finite(surv_est$surv), ]
      if (nrow(surv_est) == 0) next

      surv_est <- maybe_flip_to_survival(surv_est, pid)
      surv_est$PTID <- pid; surv_est$event <- surv_row$event[1]
      surv_est$obs_time <- surv_row$time[1]
      surv_est$group <- ifelse(surv_est$event == 1, "Converter", "Non-converter")
      curve_list[[pid]] <- surv_est
    }, error = function(e)
      cat(sprintf("    WARNING: predict failed for %s: %s\n", pid, trimws(conditionMessage(e)))))
  }

  if (length(curve_list) > 0) {
    curve_df <- do.call(rbind, curve_list)
    curve_df$PTID_label <- paste0(substr(curve_df$PTID,1,6), ifelse(curve_df$event==1,"*",""))
    n_ids    <- length(unique(curve_df$PTID))
    pal_pts  <- if (n_ids <= 12) scales::hue_pal()(n_ids)
                else colorRampPalette(c("#0072B2","#56B4E9","#D55E00","#E69F00","#009E73","#CC79A7"))(n_ids)
    vline_df <- curve_df %>% group_by(PTID_label) %>%
      summarise(obs_time = first(obs_time), event = first(event), .groups = "drop")

    p_individual <- ggplot(curve_df, aes(x=time, y=surv, colour=PTID_label, fill=PTID_label)) +
      geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.10, colour=NA) +
      geom_line(linewidth=0.9) +
      geom_vline(data=vline_df, aes(xintercept=obs_time, colour=PTID_label,
                                    linetype=factor(event)), linewidth=0.5, show.legend=FALSE) +
      scale_linetype_manual(values=c("0"="dotted","1"="dashed"), name=NULL) +
      scale_colour_manual(values=pal_pts, name="Patient (* = event)") +
      scale_fill_manual(values=pal_pts, name="Patient (* = event)") +
      scale_y_continuous(limits=c(0,1), labels=scales::percent_format(accuracy=1)) +
      facet_wrap(~group, ncol=2) +
      labs(title=paste("Joint model — individual predicted survival:", name),
           subtitle="Shaded = 95% CI  |  Vertical: event (dashed) / censor (dotted)",
           x="Time (years)", y="Predicted survival probability") +
      theme_minimal(base_size=12) +
      theme(legend.position="bottom", legend.text=element_text(size=8),
            strip.text=element_text(face="bold"))

    fname_a <- file.path(method_dir, paste0(gsub("[^A-Za-z0-9_-]","_",name), "_jm_individual_curves.png"))
    ggsave(fname_a, p_individual, width=12,
           height=max(5, ceiling(n_ids/4)*3+2), dpi=300)

    write.csv(
      curve_df[, c("PTID","PTID_label","time","surv","lower","upper","event","obs_time","group")],
      sub("\\.png$", ".csv", fname_a), row.names=FALSE
    )
    cat(sprintf("    Individual curves saved: %s\n", basename(fname_a)))
  }

  # (B) Group-averaged survival curves ----------------------------------------
  # Thesis figure: MSTE-JM_jm_survival_curves.png
  tryCatch({
    curve_file <- file.path(method_dir, paste0(gsub("[^A-Za-z0-9_-]","_",name),
                                               "_jm_individual_curves.csv"))
    if (!file.exists(curve_file)) { cat("    WARNING: individual curves CSV not found\n"); return(invisible(NULL)) }

    curve_df <- read.csv(curve_file, stringsAsFactors=FALSE) %>%
      group_by(PTID) %>%
      mutate(time_rel = time - min(time, na.rm=TRUE)) %>%
      ungroup()

    time_grid <- seq(0, max(curve_df$time_rel, na.rm=TRUE), length.out=120)
    curve_interp <- data.frame()

    for (pid in unique(curve_df$PTID)) {
      pt_data <- curve_df[curve_df$PTID == pid, ] %>% arrange(time_rel)
      grp     <- pt_data$group[1]

      surv_interp  <- approx(pt_data$time_rel, pt_data$surv,  xout=time_grid, rule=2)
      lower_interp <- approx(pt_data$time_rel, pt_data$lower, xout=time_grid, rule=2)
      upper_interp <- approx(pt_data$time_rel, pt_data$upper, xout=time_grid, rule=2)

      beyond_idx <- which(time_grid > max(pt_data$time_rel))
      if (length(beyond_idx) > 0) {
        last_2 <- tail(which(!is.na(pt_data$surv)), 2)
        if (length(last_2) == 2) {
          dt    <- pt_data$time_rel[last_2[2]] - pt_data$time_rel[last_2[1]]
          delta <- time_grid[beyond_idx] - pt_data$time_rel[last_2[2]]
          sl_s  <- (pt_data$surv[last_2[2]]  - pt_data$surv[last_2[1]])  / dt
          sl_lo <- (pt_data$lower[last_2[2]] - pt_data$lower[last_2[1]]) / dt
          sl_up <- (pt_data$upper[last_2[2]] - pt_data$upper[last_2[1]]) / dt
          last_s <- pt_data$surv[last_2[2]]; last_lo <- pt_data$lower[last_2[2]]; last_up <- pt_data$upper[last_2[2]]
          surv_interp$y[beyond_idx]  <- pmax(0, pmin(last_s,  last_s  + sl_s  * delta))
          lower_interp$y[beyond_idx] <- pmax(0, pmin(last_lo, last_lo + sl_lo * delta))
          upper_interp$y[beyond_idx] <- pmax(0, pmin(last_up, last_up + sl_up * delta))
        }
      }
      curve_interp <- rbind(curve_interp,
        data.frame(time=surv_interp$x,
                   surv  = pmin(pmax(surv_interp$y,  0), 1),
                   lower = pmin(pmax(lower_interp$y, 0), 1),
                   upper = pmin(pmax(upper_interp$y, 0), 1),
                   group = grp))
    }

    curve_agg <- curve_interp %>%
      group_by(group, time) %>%
      summarise(surv=mean(surv,na.rm=TRUE), lower=mean(lower,na.rm=TRUE),
                upper=mean(upper,na.rm=TRUE), .groups="drop")

    p_rep <- ggplot(curve_agg, aes(x=time, y=surv, colour=group, fill=group)) +
      geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2, colour=NA) +
      geom_line(aes(linetype=group), linewidth=1.2) +
      scale_colour_manual(values=c("Converter"="#D55E00","Non-converter"="#0072B2"), name="") +
      scale_fill_manual(values=c("Converter"="#D55E00","Non-converter"="#0072B2"), name="") +
      scale_linetype_manual(values=c("Converter"="dashed","Non-converter"="solid"), name="") +
      scale_y_continuous(limits=c(0,1), labels=scales::percent_format(accuracy=1)) +
      labs(title=paste("Joint model \u2014 predicted survival curves:", name),
           subtitle="Mean predicted survival per outcome group | Shaded = 95% CI",
           x="Time since landmark (years)", y="Predicted survival probability") +
      theme_minimal(base_size=14) +
      theme(legend.position="bottom", legend.text=element_text(size=12))

    fname_rep <- file.path(method_dir, paste0(gsub("[^A-Za-z0-9_-]","_",name),
                                              "_jm_survival_curves.png"))
    ggsave(fname_rep, p_rep, width=9, height=7, dpi=300)
    cat(sprintf("    Group survival curves saved: %s\n", basename(fname_rep)))

  }, error = function(e) cat(sprintf("    WARNING: Group curve plot failed: %s\n", e$message)))

  invisible(NULL)
}
```

## 2.6 Master Evaluation Function

```{r helpers-eval}
compute_all_metrics_with_figures <- function(coxFit, surv_data, long_data,
                                             jointFit = NULL, name = "Method",
                                             patient_level_data = NULL,
                                             cox_vars_hint = character(0)) {
  cat(sprintf("\n=== EVALUATING: %s ===\n", name))
  n_pts    <- length(unique(surv_data$PTID))
  n_events <- sum(surv_data$event)
  method_dir <- file.path("thesis_figures", gsub("[^A-Za-z0-9_-]","_",name))
  if (!dir.exists(method_dir)) dir.create(method_dir, recursive=TRUE)

  res <- list(
    name=name, n_patients=n_pts, n_events=n_events, event_rate=n_events/n_pts,
    cindex=NA, cindex_ci_lower=NA, cindex_ci_upper=NA,
    brier_1yr=NA, brier_3yr=NA, brier_5yr=NA,
    calibration_slope=NA, calibration_intercept=NA,
    rmst_tau=NA, rmst_low=NA, rmst_low_lower=NA, rmst_low_upper=NA,
    rmst_high=NA, rmst_high_lower=NA, rmst_high_upper=NA,
    rmst_diff=NA, rmst_diff_lower=NA, rmst_diff_upper=NA, rmst_pval=NA,
    mae_time=NA, rmse_time=NA, corr_time=NA, median_error=NA,
    n_significant=NA, km_high_median=NA, km_low_median=NA,
    jm_association_val=NA, jm_association_val_lower=NA,
    jm_association_val_upper=NA, jm_association_slope=NA,
    jm_association_slope_lower=NA, jm_association_slope_upper=NA,
    jm_dynamic_auc=NA
  )

  # 1. C-index
  res$cindex <- summary(coxFit)$concordance[1]
  cat(sprintf("C-index: %.4f\n", res$cindex))

  # 2. Cox coefficients + forest plot
  tryCatch({
    cf <- as.data.frame(summary(coxFit)$coefficients)
    cf$Variable <- rownames(cf); cf$HR <- exp(cf$coef)
    cf$HR_lower <- exp(cf$coef - 1.96*cf$`se(coef)`)
    cf$HR_upper <- exp(cf$coef + 1.96*cf$`se(coef)`)
    write.csv(cf, file.path(method_dir, paste0(gsub(" ","_",name),"_cox_coefficients.csv")), row.names=FALSE)
    res$n_significant <- sum(cf$`Pr(>|z|)` < 0.10)
    sig <- cf[cf$`Pr(>|z|)` < 0.10, ]
    if (nrow(sig) > 0 && nrow(sig) <= 25) {
      sig$HR_plot <- pmin(pmax(sig$HR,0.001),1000)
      sig$HR_lower_plot <- pmin(pmax(sig$HR_lower,0.001),1000)
      sig$HR_upper_plot <- pmin(pmax(sig$HR_upper,0.001),1000)
      fp <- ggplot(sig, aes(x=HR_plot, y=reorder(Variable,HR_plot))) +
        geom_vline(xintercept=1, linetype="dashed", color="gray50", linewidth=0.8) +
        geom_errorbarh(aes(xmin=HR_lower_plot,xmax=HR_upper_plot), height=0.2, linewidth=0.6) +
        geom_point(size=3, color="#0072B2") + scale_x_log10() +
        labs(title=paste("Hazard Ratios:",name), x="HR (log scale)", y="") +
        theme_minimal(base_size=13)
      ggsave(file.path(method_dir, paste0(gsub(" ","_",name),"_forest_plot.png")),
             fp, width=10, height=max(5,nrow(sig)*0.35), dpi=300)
    }
  }, error=function(e) cat(sprintf("WARNING (Cox coef): %s\n", e$message)))

  # 3. Baseline survival (thesis figure: MSTE-JM_baseline_survival.png)
  tryCatch({
    bsf <- basehaz(coxFit, centered=TRUE)
    bsf$surv_prob <- exp(-bsf$hazard)
    ggsave(file.path(method_dir, paste0(gsub(" ","_",name),"_baseline_survival.png")),
           ggplot(bsf, aes(time, surv_prob)) + geom_step(color="#0072B2", linewidth=1.2) +
             ylim(0,1) + labs(title=paste("Baseline Survival:",name),
                              x="Time (Years)", y="Survival Probability") + theme_minimal(base_size=13),
           width=8, height=6, dpi=300)
  }, error=function(e) cat(sprintf("WARNING (baseline): %s\n", e$message)))

  # 4. Brier score (thesis figure: MSTE-JM_brier_scores.png)
  tryCatch({
    if (n_pts < 10) stop("too few patients")
    pec_obj <- pec(list("Model"=coxFit), formula=Surv(time,event)~1,
                   data=surv_data, times=c(1,3,5), exact=FALSE,
                   cens.model="marginal", verbose=FALSE)
    bs <- pec_obj$AppErr$Model
    res$brier_1yr <- bs[1]; res$brier_3yr <- bs[2]; res$brier_5yr <- bs[3]
    cat(sprintf("Brier — 1yr: %.4f | 3yr: %.4f | 5yr: %.4f\n", bs[1], bs[2], bs[3]))
    ggsave(file.path(method_dir, paste0(gsub(" ","_",name),"_brier_scores.png")),
           ggplot(data.frame(Time=c(1,3,5), Brier=bs), aes(Time,Brier)) +
             geom_line(color="#0072B2",linewidth=1.3) + geom_point(size=4,color="#0072B2") +
             geom_hline(yintercept=0.10, linetype="dashed", color="darkgreen") +
             labs(title=paste("Brier Score:",name), x="Time (Years)", y="Brier Score") +
             theme_minimal(base_size=13),
           width=8, height=6, dpi=300)
  }, error=function(e) cat(sprintf("WARNING (Brier): %s\n", e$message)))

  # 5. KM + RMST (thesis figure: MSTE-JM_KM_stratified.png)
  tryCatch({
    surv_data$risk       <- predict(coxFit, newdata=surv_data, type="lp")
    surv_data$risk_group <- ifelse(surv_data$risk >= median(surv_data$risk), "High Risk","Low Risk")
    fs  <- survfit(Surv(time,event)~risk_group, data=surv_data)
    kmp <- ggsurvplot(fs, data=surv_data, risk.table=TRUE, pval=TRUE, conf.int=TRUE,
                      palette=c("#D55E00","#0072B2"), linetype=c("solid","dashed"),
                      legend.title="Risk Group", title=paste("Kaplan-Meier:",name),
                      xlab="Time (Years)", ylab="Progression-Free Probability")
    ggsave(file.path(method_dir, paste0(gsub(" ","_",name),"_KM_stratified.png")),
           kmp$plot, width=10, height=8, dpi=300)
    kt <- summary(fs)$table
    res$km_high_median <- kt["risk_group=High Risk","median"]
    res$km_low_median  <- kt["risk_group=Low Risk", "median"]
    cat(sprintf("KM median — High: %.2f yr | Low: %.2f yr\n", res$km_high_median, res$km_low_median))

    # RMST
    surv_data$arm <- as.numeric(surv_data$risk_group == "Low Risk")
    tau <- min(max(surv_data$time[surv_data$risk_group=="High Risk"],na.rm=TRUE),
               max(surv_data$time[surv_data$risk_group=="Low Risk"], na.rm=TRUE)) * 0.9
    rr  <- rmst2(surv_data$time, surv_data$event, surv_data$arm, tau=tau)
    res$rmst_tau  <- tau
    res$rmst_low  <- rr$RMST.arm1$rmst["Est."]
    res$rmst_high <- rr$RMST.arm0$rmst["Est."]
    res$rmst_diff <- rr$unadjusted.result[1,"Est."]
    res$rmst_pval <- rr$unadjusted.result[1,"p"]
    cat(sprintf("RMST diff: %.3f yr (p=%.4f)\n", res$rmst_diff, res$rmst_pval))
  }, error=function(e) cat(sprintf("WARNING (KM/RMST): %s\n", e$message)))

  # 6. Predicted vs observed (thesis figure: MSTE-JM_predicted_vs_observed.png)
  tryCatch({
    conv <- surv_data[surv_data$event==1,]
    if (nrow(conv) < 5) stop("too few converters")
    pred <- get_median_survival(coxFit, conv)
    obs  <- conv$time
    err  <- abs(pred - obs)
    res$mae_time <- mean(err); res$rmse_time <- sqrt(mean((pred-obs)^2))
    res$corr_time <- cor(pred, obs, method="pearson"); res$median_error <- median(err)
    cat(sprintf("MAE: %.3f yr | RMSE: %.3f yr | r: %.3f\n", res$mae_time, res$rmse_time, res$corr_time))
    pod <- data.frame(Predicted=pred, Observed=obs, PTID=conv$PTID, Error=err)
    ggsave(file.path(method_dir, paste0(gsub(" ","_",name),"_predicted_vs_observed.png")),
           ggplot(pod, aes(Observed,Predicted)) +
             geom_abline(intercept=0,slope=1,linetype="dashed",color="gray40") +
             geom_point(alpha=0.7,size=3.5,color="#0072B2") +
             geom_smooth(method="lm",se=TRUE,color="#D55E00",alpha=0.2) +
             labs(title=paste("Predicted vs Observed:",name),
                  x="Observed (Years)",y="Predicted (Years)") +
             theme_minimal(base_size=13) + coord_equal(),
           width=8, height=8, dpi=300)
    write.csv(pod, file.path(method_dir, paste0(gsub(" ","_",name),"_time_predictions.csv")), row.names=FALSE)
  }, error=function(e) cat(sprintf("WARNING (pred vs obs): %s\n", e$message)))

  # 7. Calibration
  tryCatch({
    bh  <- basehaz(coxFit, centered=FALSE)
    lp  <- predict(coxFit, newdata=surv_data, type="lp")
    t3  <- which(bh$time <= 3); if (length(t3)==0) t3 <- 1L
    H0  <- bh$hazard[max(t3)]; ps <- exp(-H0*exp(lp))
    if (diff(range(ps,na.rm=TRUE)) < 0.05) stop("range too narrow")
    nu  <- length(unique(round(ps,4)))
    ng  <- if (nu>=50) 10L else if (nu>=30) 7L else if (nu>=15) 5L else if (nu>=6) 3L else 0L
    if (ng < 3) stop("too homogeneous")
    brk <- unique(quantile(ps, seq(0,1,length.out=ng+1), na.rm=TRUE))
    if (length(brk) < 3) stop("too few breaks")
    dec <- cut(ps, breaks=brk, include.lowest=TRUE, labels=FALSE)
    cal_rows <- Filter(Negate(is.null), lapply(sort(unique(na.omit(dec))), function(g) {
      idx <- which(dec==g); if (length(idx)<5) return(NULL)
      km  <- summary(survfit(Surv(time,event)~1,data=surv_data[idx,]),times=3,extend=TRUE)
      if (length(km$surv)==0 || is.na(km$surv)) return(NULL)
      data.frame(predicted=mean(ps[idx]), observed=km$surv, n=length(idx))
    }))
    if (length(cal_rows) < 3) stop("too few groups")
    cd  <- do.call(rbind, cal_rows)
    cm  <- lm(observed~predicted, data=cd)
    res$calibration_slope     <- coef(cm)[2]
    res$calibration_intercept <- coef(cm)[1]
    cat(sprintf("Calibration — slope: %.3f | intercept: %.3f\n",
                res$calibration_slope, res$calibration_intercept))
    ggsave(file.path(method_dir, paste0(gsub("[^A-Za-z0-9_-]","_",name),"_calibration_plot.png")),
           ggplot(cd, aes(predicted,observed)) +
             geom_abline(intercept=0,slope=1,linetype="dashed",color="gray40",linewidth=0.8) +
             geom_smooth(method="lm",se=TRUE,color="#D55E00",fill="#D55E00",alpha=0.15) +
             geom_point(aes(size=n),color="#0072B2",alpha=0.85) +
             scale_size_continuous(name="Group n",range=c(3,8)) +
             xlim(0,1)+ylim(0,1)+coord_equal() +
             labs(title=paste("Calibration (3-year):",name),
                  subtitle=sprintf("Slope=%.3f | Intercept=%.3f | Groups=%d",
                                   res$calibration_slope,res$calibration_intercept,nrow(cd)),
                  x="Predicted 3-yr Survival",y="Observed 3-yr Survival") +
             theme_minimal(base_size=13),
           width=8, height=8, dpi=300)
  }, error=function(e) cat(sprintf("WARNING (calibration): %s\n", e$message)))

  # 8. Bootstrap CI (R=200)
  tryCatch({
    if (n_pts < 20) stop("too few patients")
    boot_pred_cols <- names(coef(coxFit))
    boot_formula   <- as.formula(paste("Surv(time,event) ~", paste(boot_pred_cols, collapse=" + ")))
    boot_count <- 0L
    boot_fn <- function(data, idx) {
      boot_count <<- boot_count + 1L
      if (boot_count %% 50 == 0) cat(sprintf("  Bootstrap rep %d/200...\n", boot_count))
      bd  <- data[idx, ]
      bfx <- tryCatch(coxph(boot_formula, data=bd, x=TRUE, control=coxph.control(iter.max=100)), error=function(e) NULL)
      if (is.null(bfx)) return(NA_real_)
      cv <- tryCatch(coef(bfx), error=function(e) NULL)
      if (!is.null(cv) && any(!is.finite(cv))) return(NA_real_)
      summary(bfx)$concordance[1]
    }
    br  <- boot(data=surv_data, statistic=boot_fn, R=200)
    vld <- br$t[!is.na(br$t) & br$t < 1.0]
    if (length(vld) >= 50) {
      res$cindex_ci_lower <- quantile(vld, 0.025)
      res$cindex_ci_upper <- quantile(vld, 0.975)
    } else if (length(vld) >= 10) {
      se_boot <- sd(vld)
      res$cindex_ci_lower <- res$cindex - 1.96*se_boot
      res$cindex_ci_upper <- res$cindex + 1.96*se_boot
    }
    saveRDS(br, file.path(method_dir, paste0(gsub(" ","_",name),"_bootstrap.rds")))
  }, error=function(e) cat(sprintf("WARNING (bootstrap): %s\n", e$message)))

  # 9. Joint model metrics
  if (!is.null(jointFit)) {
    jm_assoc <- extract_jm_association(jointFit, method_dir, name)
    res[names(jm_assoc)] <- jm_assoc

    res$jm_dynamic_auc <- compute_dynamic_auc(
      jointFit, long_data, surv_data, patient_level_data, cox_vars_hint, method_dir, name
    )

    plot_joint_model_survival_curves(jointFit, surv_data, long_data, method_dir, name)
  }

  # Save to master results CSV
  rdf <- as.data.frame(res, stringsAsFactors=FALSE)
  mf  <- "thesis_figures/ALL_METHODS_RESULTS.csv"
  if (file.exists(mf)) {
    ex <- read.csv(mf, stringsAsFactors=FALSE)
    ac <- union(names(ex), names(rdf))
    for (col in setdiff(ac, names(ex)))  ex[[col]]  <- NA
    for (col in setdiff(ac, names(rdf))) rdf[[col]] <- NA
    rdf <- rbind(ex[,ac], rdf[,ac])
  }
  write.csv(rdf, mf, row.names=FALSE)

  cat(sprintf("\n--- SUMMARY: %s ---\n", name))
  cat(sprintf("C-index: %.4f", res$cindex))
  if (!is.na(res$cindex_ci_lower)) cat(sprintf(" (%.4f, %.4f)", res$cindex_ci_lower, res$cindex_ci_upper))
  cat("\n")
  if (!is.na(res$brier_3yr))         cat(sprintf("Brier 3yr:  %.4f\n", res$brier_3yr))
  if (!is.na(res$calibration_slope)) cat(sprintf("Cal slope:  %.3f\n",  res$calibration_slope))
  if (!is.na(res$rmst_diff))         cat(sprintf("RMST diff:  %.3f yr (p=%.4f)\n", res$rmst_diff, res$rmst_pval))
  if (!is.na(res$mae_time))          cat(sprintf("MAE:        %.3f yr\n", res$mae_time))
  return(res)
}
```

---

# 3. Load & Prepare Data

```{r load-data}
pl_file   <- "latent_patient_level_v6.csv"
long_file <- "latent_improved_autoencoder_v6.csv"

if (!file.exists(pl_file))   stop(paste(pl_file,   "not found — check working directory"))
if (!file.exists(long_file)) stop(paste(long_file, "not found — check working directory"))

patient_level_full <- read.csv(pl_file,   stringsAsFactors=FALSE)
thesis_data_full   <- read.csv(long_file, stringsAsFactors=FALSE)
patient_level_full$PTID <- as.character(patient_level_full$PTID)
thesis_data_full$PTID   <- as.character(thesis_data_full$PTID)

if (!"split" %in% names(patient_level_full))
  stop("No 'split' column found — re-export from Python pipeline")

cat(sprintf("Patients loaded: %d\n", length(unique(patient_level_full$PTID))))
cat("Split:\n"); print(table(patient_level_full$split, useNA="always"))
```

```{r split-data}
train_pl <- patient_level_full %>% filter(split == "train")
test_pl  <- patient_level_full %>% filter(split == "val")
train_lr <- thesis_data_full   %>% filter(split == "train")
test_lr  <- thesis_data_full   %>% filter(split == "val")

cat(sprintf("Train: %d patients | Test: %d patients\n", nrow(train_pl), nrow(test_pl)))
```

---

# 4. Feature Engineering

```{r clinical-features}
# Compute per-patient MMSE slope and last MMSE from longitudinal data
compute_clinical <- function(long_df) {
  long_df %>%
    mutate(PTID=as.character(PTID), MMSE=as.numeric(MMSE), Years_bl=as.numeric(Years_bl)) %>%
    group_by(PTID) %>% arrange(Years_bl, .by_group=TRUE) %>%
    summarize(
      mmse_slope_clinical = if (n()<2 || var(Years_bl,na.rm=TRUE)<=1e-10) 0.0
                            else tryCatch(as.numeric(coef(lm(MMSE~Years_bl))[2]), error=function(e) 0.0),
      mmse_last = as.numeric(last(MMSE)),
      .groups="drop"
    ) %>% select(PTID, mmse_slope_clinical, mmse_last)
}

CLINICAL_COLS <- c("mmse_slope_clinical", "mmse_last")
clin_train    <- compute_clinical(train_lr)
clin_test     <- compute_clinical(test_lr)

# Merge clinical features into patient-level data
for (col in CLINICAL_COLS) {
  if (col %in% names(train_pl)) train_pl[[col]] <- NULL
  if (col %in% names(test_pl))  test_pl[[col]]  <- NULL
}
train_pl <- train_pl %>% left_join(clin_train, by="PTID")
test_pl  <- test_pl  %>% left_join(clin_test,  by="PTID")
```

```{r latent-features}
# Filter latent features: must have variance and not be near-zero dominated
z_final_cols <- grep("^z_final_", names(patient_level_full), value=TRUE)
z_slope_cols <- grep("^z_slope_", names(patient_level_full), value=TRUE)

keep_var <- function(cols, data)
  cols[sapply(cols, function(f) { v <- var(as.numeric(data[[f]]),na.rm=TRUE); !is.na(v) && v > 1e-10 })]
keep_bin <- function(cols, data)
  cols[sapply(cols, function(f) mean(data[[f]]==0, na.rm=TRUE) < 0.40)]

z_final_cols <- keep_bin(keep_var(z_final_cols, train_pl), train_pl)
z_slope_cols <- keep_bin(keep_var(z_slope_cols, train_pl), train_pl)
latent_pool  <- c(z_final_cols, z_slope_cols)
cat(sprintf("Latent features after filtering: %d\n", length(latent_pool)))

# Standardise all features
all_cols_m7 <- c(CLINICAL_COLS, latent_pool)
std_out     <- standardise_features(train_pl, test_pl, all_cols_m7)
train_pl    <- std_out$train
test_pl     <- std_out$test
```

```{r impute}
impute_features <- function(df, feat_cols, split_name) {
  for (col in feat_cols) {
    if (!col %in% names(df)) next
    n_na <- sum(is.na(df[[col]]))
    if (n_na > 0) {
      cat(sprintf("[%s] Imputing %d NAs in %s to 0\n", split_name, n_na, col))
      df[[col]][is.na(df[[col]])] <- 0
    }
  }
  df
}
```

---

# 5. Feature Selection

```{r feature-selection}
surv_train_sel <- train_pl %>%
  select(PTID, time_to_event, event, all_of(all_cols_m7)) %>%
  filter(!is.na(time_to_event), !is.na(event), time_to_event > 0) %>%
  rename(time = time_to_event)
surv_train_sel <- impute_features(surv_train_sel, all_cols_m7, "train-sel")
surv_train_sel <- surv_train_sel[!is.na(surv_train_sel$time) & !is.na(surv_train_sel$event), ]

n_events_m7 <- sum(surv_train_sel$event)
cat(sprintf("Training set: %d patients, %d events\n", nrow(surv_train_sel), n_events_m7))

# Select best latent feature by univariate C-index
final_features <- CLINICAL_COLS
if (length(latent_pool) > 0) {
  uni_c <- sapply(latent_pool, function(f) tryCatch(
    summary(coxph(as.formula(paste("Surv(time,event) ~", f)), data=surv_train_sel))$concordance[1],
    error=function(e) 0.5
  ))
  best_latent <- latent_pool[which.max(uni_c)]
  cat(sprintf("Best latent feature: %s (C=%.4f)\n", best_latent, max(uni_c)))
  final_features <- c(final_features, best_latent)
}
cat(sprintf("Final features: %s\n", paste(final_features, collapse=", ")))

# Best single feature for joint model Cox sub-model
uni_c_jm <- sapply(final_features, function(f) tryCatch(
  summary(coxph(as.formula(paste("Surv(time,event) ~", f)), data=surv_train_sel))$concordance[1],
  error=function(e) 0.5
))
top1_feature <- final_features[which.max(uni_c_jm)]
cat(sprintf("JM Cox feature (single): %s (C=%.4f)\n", top1_feature, max(uni_c_jm)))

latent_in_model  <- setdiff(final_features, CLINICAL_COLS)
jm_long_features <- unique(c(latent_in_model, top1_feature))
```

---

# 6. Build Training & Test Sets

```{r build-train}
train_surv <- train_pl %>%
  select(PTID, time_to_event, event, all_of(final_features)) %>%
  filter(!is.na(time_to_event), !is.na(event), time_to_event > 0) %>%
  rename(time = time_to_event)
train_surv <- impute_features(train_surv, final_features, "train-surv")
train_surv <- train_surv[!is.na(train_surv$time) & !is.na(train_surv$event), ]

train_long <- train_lr %>%
  filter(PTID %in% train_surv$PTID, !is.na(MMSE), !is.na(Years_bl)) %>%
  left_join(train_pl %>% select(PTID, any_of(jm_long_features)), by="PTID") %>%
  select(PTID, Years_bl, MMSE, any_of(jm_long_features)) %>%
  arrange(PTID, Years_bl) %>% filter(complete.cases(.))

# Align IDs; remove patients with only one visit (LME requires ≥2)
common_tr  <- intersect(train_surv$PTID, train_long$PTID)
train_surv <- train_surv %>% filter(PTID %in% common_tr)
train_long <- train_long %>% filter(PTID %in% common_tr)
single_v   <- train_long %>% group_by(PTID) %>% summarize(n=n(),.groups="drop") %>%
  filter(n < 2) %>% pull(PTID)
train_surv <- train_surv %>% filter(!PTID %in% single_v)
train_long <- train_long %>% filter(!PTID %in% single_v)

cat(sprintf("Train final: %d patients, %d events, %d visits\n",
            nrow(train_surv), sum(train_surv$event), nrow(train_long)))
```

```{r build-test}
test_surv <- test_pl %>%
  select(PTID, time_to_event, event, all_of(final_features)) %>%
  filter(!is.na(time_to_event), !is.na(event), time_to_event > 0) %>%
  rename(time = time_to_event)
test_surv <- impute_features(test_surv, final_features, "test")
test_surv <- test_surv[!is.na(test_surv$time) & !is.na(test_surv$event), ]

test_long <- test_lr %>%
  filter(PTID %in% test_surv$PTID, !is.na(MMSE), !is.na(Years_bl)) %>%
  left_join(test_pl %>% select(PTID, any_of(jm_long_features)), by="PTID") %>%
  select(PTID, Years_bl, MMSE, any_of(jm_long_features)) %>%
  arrange(PTID, Years_bl) %>% filter(complete.cases(.))

common_te <- intersect(test_surv$PTID, test_long$PTID)
test_long <- test_long %>% filter(PTID %in% common_te)

cat(sprintf("Test: %d patients, %d events\n", nrow(test_surv), sum(test_surv$event)))
cat(sprintf("Test longitudinal: %d patients, %d visits\n", length(common_te), nrow(test_long)))
```

---

# 7. Model Fitting

## 7.1 Cox Model

```{r fit-cox}
sf_m7     <- as.formula(paste("Surv(time,event) ~", paste(final_features, collapse=" + ")))
coxFit_m7 <- coxph(sf_m7, data=train_surv, x=TRUE, method="breslow")
cat(sprintf("Train C-index: %.4f\n", summary(coxFit_m7)$concordance[1]))
```

## 7.2 Linear Mixed Effects Model

```{r fit-lme}
lf_m7     <- if (length(latent_in_model) > 0)
               as.formula(paste("MMSE ~ Years_bl +", paste(latent_in_model, collapse=" + ")))
             else
               as.formula("MMSE ~ Years_bl")
lmeFit_m7 <- safe_lme(lf_m7, data=train_long)
cat("LME fitted.\n")
```

## 7.3 Joint Model

> ⏱ **This step takes ~20–40 minutes** (MCMC with 2 chains × 10,000 iterations).

```{r fit-jm}
coxFit_m7_jm  <- coxph(as.formula(paste("Surv(time,event) ~", top1_feature)),
                        data=train_surv, x=TRUE, method="breslow")
lmeFit_m7_jm  <- safe_lme(as.formula(paste("MMSE ~ Years_bl +", top1_feature)),
                           data=train_long)
jointFit_m7   <- safe_jm(coxFit_m7_jm, lmeFit_m7_jm, method_name="MAE-JM", seed=42)

if (!is.null(jointFit_m7)) cat("Joint model fitted successfully.\n") else cat("WARNING: Joint model failed.\n")
```

---

# 8. Evaluation & Figure Generation

> Figures are saved to `thesis_figures/MSTE-JM/`

```{r evaluate}
metrics_m7 <- compute_all_metrics_with_figures(
  coxFit             = coxFit_m7,
  surv_data          = test_surv,
  long_data          = test_long,
  jointFit           = jointFit_m7,
  name               = "MSTE-JM",
  patient_level_data = test_pl,
  cox_vars_hint      = top1_feature
)
```

---

# 9. Results Summary

```{r results-table}
res_df <- data.frame(
  Metric = c("C-index", "C-index 95% CI lower", "C-index 95% CI upper",
             "Brier 1yr", "Brier 3yr", "Brier 5yr",
             "Calibration slope", "RMST difference (yr)", "RMST p-value",
             "MAE (yr)", "RMSE (yr)", "Pearson r",
             "JM alpha (value)", "JM alpha (slope)", "Dynamic AUC"),
  Value = with(metrics_m7, c(
    round(cindex, 4), round(cindex_ci_lower, 4), round(cindex_ci_upper, 4),
    round(brier_1yr, 4), round(brier_3yr, 4), round(brier_5yr, 4),
    round(calibration_slope, 3), round(rmst_diff, 3), round(rmst_pval, 4),
    round(mae_time, 3), round(rmse_time, 3), round(corr_time, 3),
    round(jm_association_val, 4), round(jm_association_slope, 4),
    round(jm_dynamic_auc, 4)
  ))
)
knitr::kable(res_df, caption="MSTE-JM — Test Set Performance")
```

---

*Figures saved to: `thesis_figures/MSTE-JM/`*
