---
title: "Comparison Methods (Methods 1–6)"
author: "Tiffany"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

# Overview

This notebook runs Methods 1–6 used as baselines and ablations in the thesis.

| Method | Name | Input File(s) |
|--------|------|---------------|
| 1 | Clinical Cox (baseline) | `baseline_clinical_features.csv` |
| 2 | Image-Only CNN | `image_only_features.csv` |
| 3 | Tabular-Only Deep NN | `tabular_patient_level.csv` |
| 4 | Concatenation Fusion | `concat_patient_level.csv`, `latent_concat.csv` |
| 5 | No Autoencoder | `no_ae_patient_level.csv`, `latent_no_ae.csv` |
| 6 | AE-Only (No MTL) | `ae_only_patient_level.csv`, `latent_ae_only.csv` |

Figures are saved to `thesis_figures/<method_name>/`.

---

# 1. Setup

```{r packages}
required_packages <- c(
  "survRM2", "boot", "JMbayes2", "dplyr", "nlme",
  "survival", "pec", "survminer", "ggplot2", "gridExtra",
  "tidyr", "glmnet", "e1071", "scales"
)
for (pkg in required_packages) {
  if (!require(pkg, character.only=TRUE, quietly=TRUE)) {
    install.packages(pkg, repos="http://cran.r-project.org")
    library(pkg, character.only=TRUE)
  }
}
select    <- dplyr::select
summarize <- dplyr::summarize
if (!dir.exists("thesis_figures")) dir.create("thesis_figures")
cat("Setup complete.\n")
```

---

# 2. Shared Helpers

These are the same utilities used in `01_thesis_model.Rmd`. They must be run
before any method section below.

```{r helpers}
safe_cv_glmnet <- function(x_mat, y_mat, alpha=0.5) {
  n <- nrow(x_mat)
  if (n < 5) { cat(sprintf("  WARNING: Only %d patients\n", n)); return(NULL) }
  nfolds <- min(10L, n)
  tryCatch(
    cv.glmnet(x_mat, y_mat, family="cox", alpha=alpha, nfolds=nfolds, type.measure="C"),
    error=function(e) { cat(sprintf("  WARNING: cv.glmnet failed (%s)\n", e$message)); NULL }
  )
}

cap_features_by_epe <- function(sel, events, epe=10) {
  sel <- as.character(sel); events <- as.integer(events)
  max_f <- max(1L, floor(events/epe))
  if (length(sel) > max_f) { sel <- sel[seq_len(max_f)] }
  sel
}

standardise_features <- function(train_df, test_df, feat_cols) {
  feat_cols <- intersect(feat_cols, intersect(names(train_df), names(test_df)))
  for (f in feat_cols) {
    train_df[[f]] <- as.numeric(train_df[[f]]); test_df[[f]] <- as.numeric(test_df[[f]])
    m <- mean(train_df[[f]], na.rm=TRUE); s <- sd(train_df[[f]], na.rm=TRUE)
    if (!is.na(s) && s > 1e-10) {
      train_df[[f]] <- pmin(pmax((train_df[[f]]-m)/s,-5),5)
      test_df[[f]]  <- pmin(pmax((test_df[[f]]-m)/s,-5),5)
    } else { train_df[[f]] <- 0.0; test_df[[f]] <- 0.0 }
  }
  list(train=train_df, test=test_df)
}

select_features <- function(train_df, all_cols, n_events, method_name="") {
  all_cols <- intersect(all_cols, names(train_df))
  ok_var   <- sapply(all_cols, function(f) { v <- var(as.numeric(train_df[[f]]),na.rm=TRUE); !is.na(v) && v > 1e-10 })
  all_cols <- all_cols[ok_var]
  if (length(all_cols) == 0) return(character(0))

  x_mat  <- as.matrix(train_df[,all_cols,drop=FALSE])
  y_mat  <- Surv(train_df$time, train_df$event)
  cv_fit <- safe_cv_glmnet(x_mat, y_mat)
  sel    <- character(0)
  if (!is.null(cv_fit)) {
    coefs <- coef(cv_fit, s=cv_fit$lambda.1se); sel <- all_cols[which(coefs!=0)]
    if (length(sel) < 5) { coefs <- coef(cv_fit, s=cv_fit$lambda.min); sel <- all_cols[which(coefs!=0)] }
  }
  if (length(sel) < 5) {
    uni_c   <- sapply(all_cols, function(f) tryCatch(
      summary(coxph(as.formula(paste("Surv(time,event) ~",f)),data=train_df))$concordance[1],
      error=function(e) 0.5))
    top_idx <- order(uni_c,decreasing=TRUE)[seq_len(min(10L,length(all_cols)))]
    sel     <- all_cols[top_idx]
  }
  sel <- cap_features_by_epe(sel, n_events)
  if (length(sel) > 1) {
    test_cox <- tryCatch(
      coxph(as.formula(paste("Surv(time,event) ~",paste(sel,collapse=" + "))),
            data=train_df, x=FALSE, control=coxph.control(iter.max=30)),
      error=function(e) NULL)
    if (!is.null(test_cox)) {
      cv <- tryCatch(coef(test_cox),error=function(e) NULL)
      if (!is.null(cv) && any(!is.finite(cv))) {
        uni_c2 <- sapply(sel,function(f) tryCatch(
          summary(coxph(as.formula(paste("Surv(time,event) ~",f)),data=train_df))$concordance[1],
          error=function(e) 0.5))
        sel <- sel[which.max(uni_c2)]
      }
    }
  }
  cat(sprintf("  Selected %d features\n", length(sel))); sel
}

fit_ridge_cox <- function(train_df, sel, method_name="") {
  x_mat  <- as.matrix(train_df[,sel,drop=FALSE])
  y_mat  <- Surv(train_df$time, train_df$event)
  nfolds <- min(10L, nrow(x_mat))
  lambda_grid <- exp(seq(log(0.001),log(10),length.out=100))
  cv_ridge <- tryCatch(
    cv.glmnet(x_mat,y_mat,family="cox",alpha=0,lambda=lambda_grid,nfolds=nfolds,type.measure="C"),
    error=function(e) NULL)
  if (is.null(cv_ridge))
    return(coxph(as.formula(paste("Surv(time,event) ~",paste(sel,collapse=" + "))),
                 data=train_df, x=TRUE, method="breslow"))
  ridge_coefs <- as.numeric(coef(cv_ridge, s=cv_ridge$lambda.min)); names(ridge_coefs) <- sel
  if (max(abs(ridge_coefs)) < 0.001) {
    best_lambda <- cv_ridge$lambda[which.max(cv_ridge$cvm)]
    ridge_coefs <- as.numeric(coef(cv_ridge,s=best_lambda)); names(ridge_coefs) <- sel
  }
  tryCatch(
    coxph(as.formula(paste("Surv(time,event) ~",paste(sel,collapse=" + "))),
          data=train_df, x=TRUE, method="breslow",
          init=ridge_coefs, control=coxph.control(iter.max=0)),
    error=function(e)
      coxph(as.formula(paste("Surv(time,event) ~",paste(sel,collapse=" + "))),
            data=train_df, x=TRUE, method="breslow")
  )
}

get_median_survival <- function(coxFit, newdata) {
  bh <- basehaz(coxFit, centered=FALSE); lp <- predict(coxFit, newdata=newdata, type="lp")
  obs_cap <- quantile(newdata$time, 0.95, na.rm=TRUE)
  sapply(lp, function(l) {
    surv_i <- exp(-bh$hazard*exp(l)); idx <- which(surv_i <= 0.5)
    if (length(idx) > 0) min(bh$time[min(idx)], obs_cap)
    else min(bh$time[which.min(surv_i)], obs_cap)
  })
}

safe_lme <- function(formula, data, random_slope=TRUE) {
  ctrl <- lmeControl(opt="optim", maxIter=500, returnObject=FALSE)
  if (random_slope) {
    tryCatch(lme(formula, random=~Years_bl|PTID, data=data, control=ctrl),
             error=function(e) {
               cat(sprintf("  WARNING: Random-slope LME failed — intercept-only\n"))
               tryCatch(lme(formula, random=~1|PTID, data=data, control=ctrl),
                        error=function(e2) stop(sprintf("LME failed: %s", e2$message)))
             })
  } else {
    lme(formula, random=~1|PTID, data=data, control=ctrl)
  }
}

safe_jm <- function(cox_fit, lme_fit, time_var="Years_bl", seed=42, method_name="") {
  cat("  Fitting JM (value+slope)...\n")
  jm1 <- tryCatch(
    jm(cox_fit, lme_fit, time_var=time_var,
       functional_forms=~value(MMSE)+slope(MMSE),
       n_iter=10000, n_burnin=2000, n_thin=5, n_chains=2, seed=seed),
    error=function(e) { cat(sprintf("  WARNING: JM value+slope failed — trying value-only\n")); NULL }
  )
  if (!is.null(jm1)) { cat("  JM converged (value+slope)\n"); return(jm1) }
  jm2 <- tryCatch(
    jm(cox_fit, lme_fit, time_var=time_var,
       functional_forms=~value(MMSE),
       n_iter=7000, n_burnin=1500, n_thin=5, n_chains=2, seed=seed),
    error=function(e) { cat(sprintf("  WARNING: JM value-only also failed\n")); NULL }
  )
  if (!is.null(jm2)) cat("  JM converged (value-only)\n")
  jm2
}

extract_jm_association <- function(jointFit, method_dir, name) {
  res <- list(jm_association_val=NA, jm_association_val_lower=NA, jm_association_val_upper=NA,
              jm_association_slope=NA, jm_association_slope_lower=NA, jm_association_slope_upper=NA)
  if (is.null(jointFit)) return(res)
  sm <- tryCatch(summary(jointFit),error=function(e) NULL); if (is.null(sm)) return(res)
  tbl <- tryCatch(as.data.frame(sm$Survival),error=function(e) NULL); if (is.null(tbl)) return(res)
  cn <- colnames(tbl)
  mean_col  <- cn[grep("^[Mm]ean$|^[Ee]stimate$",cn)][1]
  lower_col <- cn[grep("2\\.5|[Ll]ower",cn)][1]
  upper_col <- cn[grep("97\\.5|[Uu]pper",cn)][1]
  rn <- rownames(tbl)
  vi <- grep("value|Assoct$",rn,ignore.case=TRUE)
  if (length(vi)>0) {
    res$jm_association_val       <- as.numeric(tbl[vi[1],mean_col])
    res$jm_association_val_lower <- as.numeric(tbl[vi[1],lower_col])
    res$jm_association_val_upper <- as.numeric(tbl[vi[1],upper_col])
  }
  si <- grep("slope|Assoct_slope",rn,ignore.case=TRUE)
  if (length(si)>0) {
    res$jm_association_slope       <- as.numeric(tbl[si[1],mean_col])
    res$jm_association_slope_lower <- as.numeric(tbl[si[1],lower_col])
    res$jm_association_slope_upper <- as.numeric(tbl[si[1],upper_col])
  }
  write.csv(tbl, file.path(method_dir, paste0(gsub(" ","_",name),"_jm_association.csv")), row.names=TRUE)
  res
}

compute_dynamic_auc <- function(jointFit, long_data, surv_data, patient_level_data,
                                cox_vars, method_dir, name, min_events_for_auc=4) {
  res_auc <- NA
  if (is.null(jointFit) || is.null(long_data) || nrow(long_data)==0) return(res_auc)
  long_data$PTID <- as.character(long_data$PTID)
  long_data$Years_bl <- as.numeric(long_data$Years_bl)
  long_data$MMSE <- as.numeric(long_data$MMSE)
  surv_data$PTID <- as.character(surv_data$PTID)
  max_obs <- max(surv_data$time,na.rm=TRUE)
  landmarks <- c(0.5,1,1.5,2)[c(0.5,1,1.5,2) < max_obs*0.80]
  cox_vars_detected <- tryCatch({
    f <- jointFit$model_info$terms_Surv_noResp
    if (!is.null(f)) setdiff(all.vars(f),c("time","event","PTID","Years_bl","MMSE")) else character(0)
  }, error=function(e) character(0))
  if (length(cox_vars_detected)==0) cox_vars_detected <- cox_vars
  surv_outcome <- surv_data %>% select(PTID,time,event)
  aucs <- c(); used_landmarks <- c()
  for (t0 in landmarks) {
    tryCatch({
      at_risk_ids <- surv_outcome$PTID[surv_outcome$time > t0]
      if (length(at_risk_ids) < 5) next
      long_t0 <- long_data %>% filter(PTID %in% at_risk_ids, Years_bl <= t0+1e-6) %>%
        left_join(surv_outcome, by="PTID")
      if (length(cox_vars_detected)>0 && !is.null(patient_level_data)) {
        missing_cox <- setdiff(cox_vars_detected, names(long_t0))
        if (length(missing_cox)>0) {
          avail <- intersect(c("PTID",missing_cox),names(patient_level_data))
          if (length(avail)>1)
            long_t0 <- long_t0 %>%
              left_join(patient_level_data %>% select(all_of(avail)) %>% distinct(PTID,.keep_all=TRUE), by="PTID")
        }
      }
      long_t0 <- long_t0[complete.cases(long_t0[,intersect(c("PTID","Years_bl","MMSE","time","event"),names(long_t0))]),]
      n_pts_t0 <- length(unique(long_t0$PTID))
      n_evt_t0 <- sum(surv_outcome$event==1 & surv_outcome$time>t0 &
                        surv_outcome$time<=(t0+1.5) & surv_outcome$PTID %in% unique(long_t0$PTID))
      if (n_pts_t0<5 || n_evt_t0<min_events_for_auc) next
      auc_obj <- tvAUC(jointFit, newdata=long_t0, Tstart=t0, Dt=1.5)
      aucs <- c(aucs, auc_obj$auc); used_landmarks <- c(used_landmarks, t0)
      cat(sprintf("    Dynamic AUC at t=%.1f: %.4f\n", t0, auc_obj$auc))
    }, error=function(e) cat(sprintf("    WARNING: tvAUC at t=%.1f failed: %s\n", t0, trimws(e$message))))
  }
  if (length(aucs)>0) {
    res_auc <- mean(aucs,na.rm=TRUE)
    write.csv(data.frame(Landmark=used_landmarks, AUC=aucs),
              file.path(method_dir, paste0(gsub(" ","_",name),"_dynamic_auc.csv")), row.names=FALSE)
  }
  res_auc
}

# Helper: read patient-level CSV and return train/test splits
build_split_surv <- function(pl_csv, z_final_pat="^z_final_", z_slope_pat="^z_slope_",
                             extra_surv_cols=character(0), method_name="") {
  if (!file.exists(pl_csv)) stop(sprintf("File not found: %s", pl_csv))
  pl <- read.csv(pl_csv, stringsAsFactors=FALSE)
  pl$PTID <- as.character(pl$PTID)
  if (!"split" %in% names(pl)) stop(sprintf("No 'split' column in %s", pl_csv))
  zf <- grep(z_final_pat, names(pl), value=TRUE)
  zs <- grep(z_slope_pat, names(pl), value=TRUE)
  all_cols <- intersect(c(zf,zs,extra_surv_cols), names(pl))
  surv_full <- pl %>%
    select(PTID, split, time_to_event, event, all_of(all_cols)) %>%
    filter(!is.na(time_to_event), !is.na(event), time_to_event>0) %>%
    rename(time=time_to_event)
  surv_full <- surv_full[complete.cases(surv_full),]
  train_raw <- surv_full %>% filter(split=="train")
  test_raw  <- surv_full %>% filter(split=="val")
  std <- standardise_features(train_raw, test_raw, all_cols)
  list(train=std$train, test=std$test, all_cols=all_cols, zf=zf, zs=zs, pl=pl)
}

# Master evaluation (metrics + figures)
compute_all_metrics_with_figures <- function(coxFit, surv_data, long_data,
                                             jointFit=NULL, name="Method",
                                             patient_level_data=NULL,
                                             cox_vars_hint=character(0)) {
  cat(sprintf("\n=== EVALUATING: %s ===\n", name))
  n_pts    <- length(unique(surv_data$PTID))
  n_events <- sum(surv_data$event)
  method_dir <- file.path("thesis_figures", gsub("[^A-Za-z0-9_-]","_",name))
  if (!dir.exists(method_dir)) dir.create(method_dir, recursive=TRUE)

  res <- list(
    name=name, n_patients=n_pts, n_events=n_events, event_rate=n_events/n_pts,
    cindex=NA, cindex_ci_lower=NA, cindex_ci_upper=NA,
    brier_1yr=NA, brier_3yr=NA, brier_5yr=NA,
    calibration_slope=NA, calibration_intercept=NA,
    rmst_tau=NA, rmst_low=NA, rmst_low_lower=NA, rmst_low_upper=NA,
    rmst_high=NA, rmst_high_lower=NA, rmst_high_upper=NA,
    rmst_diff=NA, rmst_diff_lower=NA, rmst_diff_upper=NA, rmst_pval=NA,
    mae_time=NA, rmse_time=NA, corr_time=NA, median_error=NA,
    n_significant=NA, km_high_median=NA, km_low_median=NA,
    jm_association_val=NA, jm_association_val_lower=NA, jm_association_val_upper=NA,
    jm_association_slope=NA, jm_association_slope_lower=NA, jm_association_slope_upper=NA,
    jm_dynamic_auc=NA
  )

  res$cindex <- summary(coxFit)$concordance[1]
  cat(sprintf("C-index: %.4f\n", res$cindex))

  tryCatch({
    cf <- as.data.frame(summary(coxFit)$coefficients)
    cf$Variable <- rownames(cf); cf$HR <- exp(cf$coef)
    cf$HR_lower <- exp(cf$coef-1.96*cf$`se(coef)`); cf$HR_upper <- exp(cf$coef+1.96*cf$`se(coef)`)
    write.csv(cf, file.path(method_dir,paste0(gsub(" ","_",name),"_cox_coefficients.csv")), row.names=FALSE)
    res$n_significant <- sum(cf$`Pr(>|z|)` < 0.10)
    sig <- cf[cf$`Pr(>|z|)` < 0.10,]
    if (nrow(sig)>0 && nrow(sig)<=25) {
      sig$HR_plot <- pmin(pmax(sig$HR,0.001),1000)
      sig$HR_lower_plot <- pmin(pmax(sig$HR_lower,0.001),1000)
      sig$HR_upper_plot <- pmin(pmax(sig$HR_upper,0.001),1000)
      fp <- ggplot(sig,aes(x=HR_plot,y=reorder(Variable,HR_plot))) +
        geom_vline(xintercept=1,linetype="dashed",color="gray50",linewidth=0.8) +
        geom_errorbarh(aes(xmin=HR_lower_plot,xmax=HR_upper_plot),height=0.2,linewidth=0.6) +
        geom_point(size=3,color="#0072B2") + scale_x_log10() +
        labs(title=paste("Hazard Ratios:",name),x="HR (log scale)",y="") + theme_minimal(base_size=13)
      ggsave(file.path(method_dir,paste0(gsub(" ","_",name),"_forest_plot.png")),
             fp, width=10, height=max(5,nrow(sig)*0.35), dpi=300)
    }
  }, error=function(e) cat(sprintf("WARNING (Cox coef): %s\n",e$message)))

  tryCatch({
    bsf <- basehaz(coxFit,centered=TRUE); bsf$surv_prob <- exp(-bsf$hazard)
    ggsave(file.path(method_dir,paste0(gsub(" ","_",name),"_baseline_survival.png")),
           ggplot(bsf,aes(time,surv_prob))+geom_step(color="#0072B2",linewidth=1.2)+ylim(0,1)+
             labs(title=paste("Baseline Survival:",name),x="Time (Years)",y="Survival Prob.")+theme_minimal(base_size=13),
           width=8,height=6,dpi=300)
  }, error=function(e) cat(sprintf("WARNING (baseline): %s\n",e$message)))

  tryCatch({
    if (n_pts<10) stop("too few patients")
    pec_obj <- pec(list("Model"=coxFit),formula=Surv(time,event)~1,data=surv_data,
                   times=c(1,3,5),exact=FALSE,cens.model="marginal",verbose=FALSE)
    bs <- pec_obj$AppErr$Model
    res$brier_1yr <- bs[1]; res$brier_3yr <- bs[2]; res$brier_5yr <- bs[3]
    cat(sprintf("Brier — 1yr: %.4f | 3yr: %.4f | 5yr: %.4f\n",bs[1],bs[2],bs[3]))
    ggsave(file.path(method_dir,paste0(gsub(" ","_",name),"_brier_scores.png")),
           ggplot(data.frame(Time=c(1,3,5),Brier=bs),aes(Time,Brier))+
             geom_line(color="#0072B2",linewidth=1.3)+geom_point(size=4,color="#0072B2")+
             geom_hline(yintercept=0.10,linetype="dashed",color="darkgreen")+
             labs(title=paste("Brier Score:",name),x="Time (Years)",y="Brier Score")+theme_minimal(base_size=13),
           width=8,height=6,dpi=300)
  }, error=function(e) cat(sprintf("WARNING (Brier): %s\n",e$message)))

  tryCatch({
    surv_data$risk <- predict(coxFit,newdata=surv_data,type="lp")
    surv_data$risk_group <- ifelse(surv_data$risk>=median(surv_data$risk),"High Risk","Low Risk")
    fs  <- survfit(Surv(time,event)~risk_group,data=surv_data)
    kmp <- ggsurvplot(fs,data=surv_data,risk.table=TRUE,pval=TRUE,conf.int=TRUE,
                      palette=c("#D55E00","#0072B2"),linetype=c("solid","dashed"),
                      legend.title="Risk Group",title=paste("Kaplan-Meier:",name),
                      xlab="Time (Years)",ylab="Progression-Free Probability")
    ggsave(file.path(method_dir,paste0(gsub(" ","_",name),"_KM_stratified.png")),
           kmp$plot,width=10,height=8,dpi=300)
    kt <- summary(fs)$table
    res$km_high_median <- kt["risk_group=High Risk","median"]
    res$km_low_median  <- kt["risk_group=Low Risk","median"]
    cat(sprintf("KM median — High: %.2f yr | Low: %.2f yr\n",res$km_high_median,res$km_low_median))
    surv_data$arm <- as.numeric(surv_data$risk_group=="Low Risk")
    tau <- min(max(surv_data$time[surv_data$risk_group=="High Risk"],na.rm=TRUE),
               max(surv_data$time[surv_data$risk_group=="Low Risk"],na.rm=TRUE)) * 0.9
    rr  <- rmst2(surv_data$time,surv_data$event,surv_data$arm,tau=tau)
    res$rmst_tau  <- tau; res$rmst_low <- rr$RMST.arm1$rmst["Est."]
    res$rmst_high <- rr$RMST.arm0$rmst["Est."]; res$rmst_diff <- rr$unadjusted.result[1,"Est."]
    res$rmst_pval <- rr$unadjusted.result[1,"p"]
    cat(sprintf("RMST diff: %.3f yr (p=%.4f)\n",res$rmst_diff,res$rmst_pval))
  }, error=function(e) cat(sprintf("WARNING (KM/RMST): %s\n",e$message)))

  tryCatch({
    conv <- surv_data[surv_data$event==1,]; if (nrow(conv)<5) stop("too few converters")
    pred <- get_median_survival(coxFit,conv); obs <- conv$time; err <- abs(pred-obs)
    res$mae_time <- mean(err); res$rmse_time <- sqrt(mean((pred-obs)^2))
    res$corr_time <- cor(pred,obs,method="pearson"); res$median_error <- median(err)
    cat(sprintf("MAE: %.3f yr | RMSE: %.3f yr | r: %.3f\n",res$mae_time,res$rmse_time,res$corr_time))
    pod <- data.frame(Predicted=pred,Observed=obs,PTID=conv$PTID,Error=err)
    ggsave(file.path(method_dir,paste0(gsub(" ","_",name),"_predicted_vs_observed.png")),
           ggplot(pod,aes(Observed,Predicted))+
             geom_abline(intercept=0,slope=1,linetype="dashed",color="gray40")+
             geom_point(alpha=0.7,size=3.5,color="#0072B2")+
             geom_smooth(method="lm",se=TRUE,color="#D55E00",alpha=0.2)+
             labs(title=paste("Predicted vs Observed:",name),x="Observed (Years)",y="Predicted (Years)")+
             theme_minimal(base_size=13)+coord_equal(),
           width=8,height=8,dpi=300)
  }, error=function(e) cat(sprintf("WARNING (pred vs obs): %s\n",e$message)))

  tryCatch({
    if (n_pts<20) stop("too few patients")
    boot_pred_cols <- names(coef(coxFit))
    boot_formula   <- as.formula(paste("Surv(time,event) ~",paste(boot_pred_cols,collapse=" + ")))
    boot_count <- 0L
    boot_fn <- function(data,idx) {
      boot_count <<- boot_count+1L
      if (boot_count %% 50==0) cat(sprintf("  Bootstrap rep %d/200...\n",boot_count))
      bd <- data[idx,]
      bfx <- tryCatch(coxph(boot_formula,data=bd,x=TRUE,control=coxph.control(iter.max=100)),error=function(e) NULL)
      if (is.null(bfx)) return(NA_real_)
      cv <- tryCatch(coef(bfx),error=function(e) NULL)
      if (!is.null(cv) && any(!is.finite(cv))) return(NA_real_)
      summary(bfx)$concordance[1]
    }
    br  <- boot(data=surv_data,statistic=boot_fn,R=200)
    vld <- br$t[!is.na(br$t) & br$t<1.0]
    if (length(vld)>=50) {
      res$cindex_ci_lower <- quantile(vld,0.025); res$cindex_ci_upper <- quantile(vld,0.975)
    } else if (length(vld)>=10) {
      se_boot <- sd(vld)
      res$cindex_ci_lower <- res$cindex-1.96*se_boot; res$cindex_ci_upper <- res$cindex+1.96*se_boot
    }
    saveRDS(br, file.path(method_dir,paste0(gsub(" ","_",name),"_bootstrap.rds")))
  }, error=function(e) cat(sprintf("WARNING (bootstrap): %s\n",e$message)))

  if (!is.null(jointFit)) {
    jm_assoc <- extract_jm_association(jointFit, method_dir, name)
    res[names(jm_assoc)] <- jm_assoc
    res$jm_dynamic_auc <- compute_dynamic_auc(
      jointFit, long_data, surv_data, patient_level_data, cox_vars_hint, method_dir, name)
  }

  rdf <- as.data.frame(res, stringsAsFactors=FALSE)
  mf  <- "thesis_figures/ALL_METHODS_RESULTS.csv"
  if (file.exists(mf)) {
    ex <- read.csv(mf,stringsAsFactors=FALSE)
    ac <- union(names(ex),names(rdf))
    for (col in setdiff(ac,names(ex)))  ex[[col]]  <- NA
    for (col in setdiff(ac,names(rdf))) rdf[[col]] <- NA
    rdf <- rbind(ex[,ac],rdf[,ac])
  }
  write.csv(rdf,mf,row.names=FALSE)

  cat(sprintf("\n--- SUMMARY: %s ---\n", name))
  cat(sprintf("C-index: %.4f", res$cindex))
  if (!is.na(res$cindex_ci_lower)) cat(sprintf(" (%.4f, %.4f)",res$cindex_ci_lower,res$cindex_ci_upper))
  cat("\n")
  if (!is.na(res$brier_3yr))         cat(sprintf("Brier 3yr:  %.4f\n",res$brier_3yr))
  if (!is.na(res$calibration_slope)) cat(sprintf("Cal slope:  %.3f\n",res$calibration_slope))
  if (!is.na(res$rmst_diff))         cat(sprintf("RMST diff:  %.3f yr (p=%.4f)\n",res$rmst_diff,res$rmst_pval))
  if (!is.na(res$mae_time))          cat(sprintf("MAE:        %.3f yr\n",res$mae_time))
  return(res)
}
```

---

# 3. Method 1: Clinical Cox (Baseline)

> **Input:** `baseline_clinical_features.csv`
> **Features:** AGE, PTGENDER, PTEDUCAT, ADAS13 + longitudinal MMSE (joint model)

```{r method1}
metrics_baseline <- NULL
tryCatch({
  if (!file.exists("baseline_clinical_features.csv")) stop("File not found")
  dat <- read.csv("baseline_clinical_features.csv", stringsAsFactors=FALSE)
  dat$PTID <- as.character(dat$PTID)
  if (!"split" %in% names(dat)) stop("No 'split' column")

  long_full <- dat %>%
    filter(!is.na(MMSE),!is.na(Years_bl)) %>%
    select(PTID,split,Years_bl,MMSE,ADAS13,AGE,PTGENDER,PTEDUCAT) %>%
    arrange(PTID,Years_bl) %>% filter(complete.cases(.))

  surv_full <- dat %>%
    group_by(PTID) %>%
    summarize(time=first(time_to_event),event=first(event),AGE=first(AGE),
              PTGENDER=first(PTGENDER),PTEDUCAT=first(PTEDUCAT),ADAS13=first(ADAS13),
              split=first(split),.groups="drop") %>% filter(complete.cases(.))

  surv_train <- surv_full %>% filter(split=="train")
  surv_test  <- surv_full %>% filter(split=="val")
  long_train <- long_full %>% filter(PTID %in% surv_train$PTID)
  long_test  <- long_full %>% filter(PTID %in% surv_test$PTID)

  # Remove patients with only one visit
  single_v <- long_train %>% group_by(PTID) %>% filter(n()<2) %>% pull(PTID) %>% unique()
  long_train <- long_train %>% filter(!PTID %in% single_v)
  surv_train <- surv_train %>% filter(PTID %in% intersect(surv_train$PTID, long_train$PTID))
  long_train <- long_train %>% filter(PTID %in% surv_train$PTID)

  cat(sprintf("Train: %d patients, %d events\n", nrow(surv_train), sum(surv_train$event)))
  cat(sprintf("Test:  %d patients, %d events\n", nrow(surv_test),  sum(surv_test$event)))

  lmeFit_baseline <- safe_lme(MMSE~Years_bl+AGE+PTGENDER+PTEDUCAT+ADAS13, data=long_train)
  coxFit_baseline <- coxph(Surv(time,event)~AGE+PTGENDER+PTEDUCAT+ADAS13,
                           data=surv_train, x=TRUE, ties="breslow")
  cat(sprintf("Train C-index: %.4f\n", summary(coxFit_baseline)$concordance[1]))

  jointFit_baseline <- safe_jm(coxFit_baseline, lmeFit_baseline, method_name="Clinical Cox")

  metrics_baseline <- compute_all_metrics_with_figures(
    coxFit_baseline, surv_test, long_test, jointFit_baseline,
    "Clinical Cox", patient_level_data=surv_test
  )
}, error=function(e) cat(sprintf("ERROR in Method 1: %s\n", e$message)))
```

---

# 4. Method 2: Image-Only CNN

> **Input:** `image_only_features.csv`
> **Features:** CNN image embeddings (`img_feat_*` or `z_*`)

```{r method2}
metrics_image <- NULL
tryCatch({
  if (!file.exists("image_only_features.csv")) stop("File not found")
  dat <- read.csv("image_only_features.csv", stringsAsFactors=FALSE)
  dat$PTID <- as.character(dat$PTID)
  img_features <- grep("^img_feat_|^z_", names(dat), value=TRUE)
  if (length(img_features)==0) stop("No image features found")

  surv_full <- dat %>%
    group_by(PTID) %>%
    summarize(time=first(time_to_event),event=first(event),split=first(split),
              across(all_of(img_features),first),.groups="drop") %>% filter(complete.cases(.))

  train_raw <- surv_full %>% filter(split=="train")
  test_raw  <- surv_full %>% filter(split=="val")
  std <- standardise_features(train_raw, test_raw, img_features)
  surv_train <- std$train; surv_test <- std$test

  if (length(img_features) > 10) {
    img_vars <- sapply(surv_train[img_features], var, na.rm=TRUE)
    img_features <- img_features[order(img_vars,decreasing=TRUE)[1:10]]
  }
  img_features <- cap_features_by_epe(img_features, sum(surv_train$event))

  cat(sprintf("Train: %d patients, %d events\n", nrow(surv_train), sum(surv_train$event)))
  coxFit_image <- fit_ridge_cox(surv_train, img_features, "Image-Only")
  cat(sprintf("Train C-index: %.4f\n", summary(coxFit_image)$concordance[1]))

  metrics_image <- compute_all_metrics_with_figures(
    coxFit_image, surv_test %>% select(PTID,time,event,all_of(img_features)),
    NULL, NULL, "Image-Only CNN"
  )
}, error=function(e) cat(sprintf("ERROR in Method 2: %s\n", e$message)))
```

---

# 5. Method 3: Tabular-Only Deep NN

> **Input:** `tabular_patient_level.csv`

```{r method3}
metrics_tabular <- NULL
tryCatch({
  out <- build_split_surv("tabular_patient_level.csv", method_name="Tabular")
  tab_train <- out$train; tab_test <- out$test; all_cols <- out$all_cols
  cat(sprintf("Train: %d patients, %d events\n", nrow(tab_train), sum(tab_train$event)))

  sel <- select_features(tab_train, all_cols, sum(tab_train$event), "Tabular-Only")
  coxFit_tabular <- fit_ridge_cox(tab_train, sel, "Tabular-Only")
  cat(sprintf("Train C-index: %.4f\n", summary(coxFit_tabular)$concordance[1]))

  metrics_tabular <- compute_all_metrics_with_figures(
    coxFit_tabular, tab_test %>% select(PTID,time,event,all_of(sel)),
    NULL, NULL, "Tabular-Only NN"
  )
}, error=function(e) cat(sprintf("ERROR in Method 3: %s\n", e$message)))
```

---

# 6. Method 4: Concatenation Fusion

> **Input:** `concat_patient_level.csv`, `latent_concat.csv`

```{r method4}
metrics_concat <- NULL
tryCatch({
  long_csv <- read.csv("latent_concat.csv", stringsAsFactors=FALSE)
  long_csv$PTID <- as.character(long_csv$PTID)
  out <- build_split_surv("concat_patient_level.csv", method_name="Concat")
  concat_train <- out$train; concat_test <- out$test; all_cols <- out$all_cols

  cat(sprintf("Train: %d patients, %d events\n", nrow(concat_train), sum(concat_train$event)))
  sel <- select_features(concat_train, all_cols, sum(concat_train$event), "Concat")
  coxFit_concat <- fit_ridge_cox(concat_train, sel, "Concat")

  latent_vals_train <- concat_train %>% select(PTID, all_of(sel))
  long_train <- long_csv %>%
    filter(PTID %in% concat_train$PTID, !is.na(MMSE), !is.na(Years_bl)) %>%
    left_join(latent_vals_train, by="PTID") %>%
    select(PTID,Years_bl,MMSE,all_of(sel)) %>% arrange(PTID,Years_bl) %>% filter(complete.cases(.))

  single_v <- long_train %>% group_by(PTID) %>% filter(n()<2) %>% pull(PTID) %>% unique()
  long_train   <- long_train   %>% filter(!PTID %in% single_v)
  concat_train <- concat_train %>% filter(PTID %in% intersect(concat_train$PTID, long_train$PTID))
  long_train   <- long_train   %>% filter(PTID %in% concat_train$PTID)

  lmeFit_concat <- safe_lme(as.formula(paste("MMSE ~ Years_bl +",paste(sel,collapse=" + "))), data=long_train)

  uni_c_jm <- sapply(sel, function(f) tryCatch(
    summary(coxph(as.formula(paste("Surv(time,event) ~",f)),data=concat_train))$concordance[1],
    error=function(e) 0.5))
  top1_concat <- sel[which.max(uni_c_jm)]
  coxFit_concat_jm <- coxph(as.formula(paste("Surv(time,event) ~",top1_concat)),
                            data=concat_train, x=TRUE, method="breslow")
  lmeFit_concat_jm <- safe_lme(as.formula(paste("MMSE ~ Years_bl +",top1_concat)), data=long_train)
  jointFit_concat  <- safe_jm(coxFit_concat_jm, lmeFit_concat_jm, method_name="Concat", seed=42)

  latent_vals_test <- concat_test %>% select(PTID, all_of(sel))
  long_test <- long_csv %>%
    filter(PTID %in% concat_test$PTID, !is.na(MMSE), !is.na(Years_bl)) %>%
    left_join(latent_vals_test, by="PTID") %>%
    select(PTID,Years_bl,MMSE,all_of(sel)) %>% arrange(PTID,Years_bl) %>% filter(complete.cases(.))

  metrics_concat <- compute_all_metrics_with_figures(
    coxFit_concat, concat_test %>% select(PTID,time,event,all_of(sel)),
    long_test, jointFit_concat, "Concatenation Fusion",
    patient_level_data=concat_test, cox_vars_hint=top1_concat
  )
}, error=function(e) cat(sprintf("ERROR in Method 4: %s\n", e$message)))
```

---

# 7. Method 5: No Autoencoder

> **Input:** `no_ae_patient_level.csv`, `latent_no_ae.csv`

```{r method5}
metrics_noae <- NULL
tryCatch({
  long_csv <- read.csv("latent_no_ae.csv", stringsAsFactors=FALSE)
  long_csv$PTID <- as.character(long_csv$PTID)
  out <- build_split_surv("no_ae_patient_level.csv", method_name="No-AE")
  noae_train <- out$train; noae_test <- out$test; all_cols <- out$all_cols

  cat(sprintf("Train: %d patients, %d events\n", nrow(noae_train), sum(noae_train$event)))
  sel <- select_features(noae_train, all_cols, sum(noae_train$event), "No-AE")
  coxFit_noae <- fit_ridge_cox(noae_train, sel, "No-AE")

  latent_vals_train <- noae_train %>% select(PTID, all_of(sel))
  long_train <- long_csv %>%
    filter(PTID %in% noae_train$PTID, !is.na(MMSE), !is.na(Years_bl)) %>%
    left_join(latent_vals_train, by="PTID") %>%
    select(PTID,Years_bl,MMSE,all_of(sel)) %>% arrange(PTID,Years_bl) %>% filter(complete.cases(.))

  single_v   <- long_train %>% group_by(PTID) %>% filter(n()<2) %>% pull(PTID) %>% unique()
  long_train <- long_train %>% filter(!PTID %in% single_v)
  noae_train <- noae_train %>% filter(PTID %in% intersect(noae_train$PTID, long_train$PTID))
  long_train <- long_train %>% filter(PTID %in% noae_train$PTID)

  lmeFit_noae <- safe_lme(as.formula(paste("MMSE ~ Years_bl +",paste(sel,collapse=" + "))), data=long_train)

  uni_c_jm <- sapply(sel, function(f) tryCatch(
    summary(coxph(as.formula(paste("Surv(time,event) ~",f)),data=noae_train))$concordance[1],
    error=function(e) 0.5))
  top1_noae <- sel[which.max(uni_c_jm)]
  coxFit_noae_jm <- coxph(as.formula(paste("Surv(time,event) ~",top1_noae)),
                          data=noae_train, x=TRUE, method="breslow")
  lmeFit_noae_jm <- safe_lme(as.formula(paste("MMSE ~ Years_bl +",top1_noae)), data=long_train)
  jointFit_noae  <- safe_jm(coxFit_noae_jm, lmeFit_noae_jm, method_name="No-AE", seed=42)

  latent_vals_test <- noae_test %>% select(PTID, all_of(sel))
  long_test <- long_csv %>%
    filter(PTID %in% noae_test$PTID, !is.na(MMSE), !is.na(Years_bl)) %>%
    left_join(latent_vals_test, by="PTID") %>%
    select(PTID,Years_bl,MMSE,all_of(sel)) %>% arrange(PTID,Years_bl) %>% filter(complete.cases(.))

  metrics_noae <- compute_all_metrics_with_figures(
    coxFit_noae, noae_test %>% select(PTID,time,event,all_of(sel)),
    long_test, jointFit_noae, "No Autoencoder",
    patient_level_data=noae_test, cox_vars_hint=top1_noae
  )
}, error=function(e) cat(sprintf("ERROR in Method 5: %s\n", e$message)))
```

---

# 8. Method 6: AE-Only (No MTL)

> **Input:** `ae_only_patient_level.csv`, `latent_ae_only.csv`
> **Note:** AE-only has no survival signal in training — a lower C-index is expected.

```{r method6}
metrics_ae <- NULL
tryCatch({
  long_csv <- read.csv("latent_ae_only.csv", stringsAsFactors=FALSE)
  long_csv$PTID <- as.character(long_csv$PTID)
  out <- build_split_surv("ae_only_patient_level.csv", method_name="AE-Only")
  ae_train <- out$train; ae_test <- out$test; all_cols <- out$all_cols

  cat(sprintf("Train: %d patients, %d events\n", nrow(ae_train), sum(ae_train$event)))
  sel <- select_features(ae_train, all_cols, sum(ae_train$event), "AE-Only")
  coxFit_ae <- fit_ridge_cox(ae_train, sel, "AE-Only")

  latent_vals_train <- ae_train %>% select(PTID, all_of(sel))
  long_train <- long_csv %>%
    filter(PTID %in% ae_train$PTID, !is.na(MMSE), !is.na(Years_bl)) %>%
    left_join(latent_vals_train, by="PTID") %>%
    select(PTID,Years_bl,MMSE,all_of(sel)) %>% arrange(PTID,Years_bl) %>% filter(complete.cases(.))

  single_v <- long_train %>% group_by(PTID) %>% filter(n()<2) %>% pull(PTID) %>% unique()
  long_train <- long_train %>% filter(!PTID %in% single_v)
  ae_train   <- ae_train   %>% filter(PTID %in% intersect(ae_train$PTID, long_train$PTID))
  long_train <- long_train %>% filter(PTID %in% ae_train$PTID)

  lmeFit_ae <- safe_lme(as.formula(paste("MMSE ~ Years_bl +",paste(sel,collapse=" + "))), data=long_train)

  uni_c_jm <- sapply(sel, function(f) tryCatch(
    summary(coxph(as.formula(paste("Surv(time,event) ~",f)),data=ae_train))$concordance[1],
    error=function(e) 0.5))
  top1_ae <- sel[which.max(uni_c_jm)]
  coxFit_ae_jm <- coxph(as.formula(paste("Surv(time,event) ~",top1_ae)),
                        data=ae_train, x=TRUE, method="breslow")
  lmeFit_ae_jm <- safe_lme(as.formula(paste("MMSE ~ Years_bl +",top1_ae)), data=long_train)
  jointFit_ae  <- safe_jm(coxFit_ae_jm, lmeFit_ae_jm, method_name="AE-Only", seed=42)

  latent_vals_test <- ae_test %>% select(PTID, all_of(sel))
  long_test <- long_csv %>%
    filter(PTID %in% ae_test$PTID, !is.na(MMSE), !is.na(Years_bl)) %>%
    left_join(latent_vals_test, by="PTID") %>%
    select(PTID,Years_bl,MMSE,all_of(sel)) %>% arrange(PTID,Years_bl) %>% filter(complete.cases(.))

  metrics_ae <- compute_all_metrics_with_figures(
    coxFit_ae, ae_test %>% select(PTID,time,event,all_of(sel)),
    long_test, jointFit_ae, "AE-Only (No MTL)",
    patient_level_data=ae_test, cox_vars_hint=top1_ae
  )
}, error=function(e) cat(sprintf("ERROR in Method 6: %s\n", e$message)))
```

---

# 9. Combined Results Table

```{r combined-results}
all_results_file <- "thesis_figures/ALL_METHODS_RESULTS.csv"
if (file.exists(all_results_file)) {
  all_res <- read.csv(all_results_file, stringsAsFactors=FALSE)
  display_cols <- c("name","n_patients","n_events",
                    "cindex","cindex_ci_lower","cindex_ci_upper",
                    "brier_3yr","calibration_slope",
                    "rmst_diff","rmst_pval","mae_time","jm_dynamic_auc")
  display_cols <- intersect(display_cols, names(all_res))
  knitr::kable(all_res[, display_cols],
               digits=3,
               caption="All Methods — Test Set Performance",
               row.names=FALSE)
} else {
  cat("Results file not yet created. Run all methods above first.\n")
}
```

---

*Figures saved to: `thesis_figures/<method_name>/`*
*Combined results: `thesis_figures/ALL_METHODS_RESULTS.csv`*
